---
title: "R Notebook"
output: html_notebook
---

```{r}
library(dplyr)
library(ggplot2)
library(lme4)
library(mgcv)
library(EnvStats)
```


Data Source:
Used in Udacity course “intro to data science” Udacity course.
New York City subway passenger statistics, with the weather variables. 
https://www.dropbox.com/s/1lpoeh2w6px4diu/improved-dataset.zip?dl=0

```{r}
df <- read.csv('improved-dataset/improved-dataset/turnstile_weather_v2.csv')
head(df,200)
```


2.3.1. UNIT vs. station

```{r}
df %>% group_by(UNIT,station) %>% count()
```

2.3.2. Deep dive in UNIT

```{r}
df %>% select(UNIT) %>% unique()
```

```{r}
df %>% group_by(UNIT) %>% summarize(mean.per.unit = mean(ENTRIESn_hourly)) %>%
  ggplot(aes(mean.per.unit)) + geom_histogram() + labs(x='Average Entry per UNIT')
```

```{r}
df %>% group_by(UNIT) %>% summarize(mean.log.per.unit = mean(log(ENTRIESn_hourly+1))) %>%
  ggplot(aes(mean.log.per.unit)) + geom_histogram() + labs(x='Average log(Entry+1) per UNIT')
```

```{r}
df %>% group_by(UNIT) %>% summarize(mean.log.per.unit = mean(log(ENTRIESn_hourly+1))) %>%
  filter(mean.log.per.unit>=3) %>%
  ggplot(aes(mean.log.per.unit)) + geom_histogram() + labs(x='Average log(Entry+1) per UNIT')
```


2.3.3. ENTRIESn_hourly vs. DATEn

```{r}
df %>% ggplot(aes(x=DATEn,y=ENTRIESn_hourly)) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90))
```

2.3.4. ENTRIESn_hourly vs. hour

```{r}
df1 <- df %>% 
  filter(weekday==1)
df1 %>%
  ggplot(aes(x=as.factor(hour),y=ENTRIESn_hourly)) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90))
```

2.3.5. Summary of data transformation and filtering
```{r}
df2 <- df1 %>% filter(hour==16) %>% filter(ENTRIESn_hourly>=exp(3)-1)
```

```{r}
df2 %>% group_by(UNIT) %>% count()
df2 %>% group_by(fog) %>% count()
df2 %>% group_by(precipi) %>% count()
df2 %>% group_by(rain) %>% count()
df2 %>% group_by(tempi) %>% count()
df2 %>% group_by(wspdi) %>% count()
```

```{r}
df_reg <- df2 %>% 
  mutate(log.ent=log(ENTRIESn_hourly+1)) %>% 
  mutate(rain.tempi = rain*tempi, rain.wspdi=rain*wspdi) %>% 
  select(log.ent,UNIT,fog,rain,tempi,wspdi, rain.tempi, rain.wspdi,ENTRIESn_hourly)
```

```{r}
library(caret)
set.seed(1000)
train.index <- createDataPartition(df_reg$UNIT, p = .8, list = FALSE)
df_reg.train <- df_reg[ train.index,]
df_reg.test  <- df_reg[-train.index,]
```

```{r}
summary(df_reg)
```

```{r}
length(unique(df_reg$UNIT))
```


2.4.
```{r}
rsq <- function(actual,preds) {
  rss <- sum((preds - actual) ^ 2)  ## residual sum of squares
  tss <- sum((actual - mean(actual)) ^ 2)  ## total sum of squares
  rsq <- 1 - rss/tss
}
```

```{r}
draw_qqplot.train <- function(model) {
  qqnorm(df_reg.train[,1] - predict(model,df_reg.train[,-1]), pch = 1, frame = FALSE)
  qqline(df_reg.train[,1] - predict(model,df_reg.train[,-1]), col = "steelblue", lwd = 2) 
}

draw_qqplot.test <- function(model) {
  qqnorm(df_reg.test[,1] - predict(model,df_reg.test[,-1]), pch = 1, frame = FALSE)
  qqline(df_reg.test[,1] - predict(model,df_reg.test[,-1]), col = "steelblue", lwd = 2) 
}
```




2.4.1. All-in as fixed effects variables
```{r}
m1 <- bam(log.ent~fog+rain+tempi+wspdi+as.factor(UNIT),data=df_reg.train)
summary(m1)
draw_qqplot.train(m1)
plot(df_reg.train$log.ent,df_reg.train$log.ent - fitted(m1,df_reg.train))
rsq1 <- rsq(df_reg.train$log.ent,predict(m1,df_reg.train[,-1]))
rsq1
rsq1 <- rsq(df_reg.test$log.ent,predict(m1,df_reg.test,type='response'))
rsq1
```
```{r}
AIC(m1)
```

```{r}
plot(df_reg.train$log.ent,residuals(m1,type='deviance'))
```


```{r}
#m1 <- lm(log.ent~fog+rain+tempi+wspdi+as.factor(UNIT),data=df_reg.train)
#summary(m1)
#plot(m1)
#draw_qqplot.test(m1)
#rsq1 <- rsq(df_reg.test$log.ent,predict(m1,df_reg.test[,-1]))
#rsq1
```


2.4.2. Smooth term on tempi and wspdi
```{r}
m2 <- bam(log.ent~fog+rain+s(tempi)+s(wspdi)+as.factor(UNIT),data=df_reg.train)
plot(m2)
summary(m2)
draw_qqplot.train(m2)
plot(df_reg.train$log.ent,df_reg.train$log.ent - fitted(m2,df_reg.train))
rsq2 <- rsq(df_reg.train$log.ent,predict(m2,df_reg.train[,-1]))
rsq2
rsq2 <- rsq(df_reg.test$log.ent,predict(m2,df_reg.test,type='response'))
rsq2
```

```{r}
AIC(m2)
```


2.4.3.Interaction term
```{r}
m3 <- bam(log.ent~fog+rain+tempi+wspdi+rain.tempi+rain.wspdi+as.factor(UNIT),
          select=TRUE,
          data=df_reg.train)
#plot(m3)
summary(m3)
draw_qqplot.train(m3)
plot(df_reg.train$log.ent,df_reg.train$log.ent - fitted(m3,df_reg.train))
rsq3 <- rsq(df_reg.train$log.ent,predict(m3,df_reg.train[,-1]))
rsq3
rsq3 <- rsq(df_reg.test$log.ent,predict(m3,df_reg.test,type='response'))
rsq3
```

```{r}
AIC(m3)
```

2.4.4. Random Effect on UNIT
```{r}
m4 <- gamm(log.ent~fog+rain+s(tempi)+s(wspdi),random=list(UNIT=~1),data=df_reg.train)
m4
rsq4 <- rsq(df_reg.train$log.ent,fitted(m4$lme))
rsq4
plot(df_reg.train$log.ent,df_reg.train[,1] - fitted(m4$lme))
qqnorm(df_reg.train[,1] - fitted(m4$lme), pch = 1, frame = FALSE)
qqline(df_reg.train[,1] - fitted(m4$lme), col = "steelblue", lwd = 2) 
```


```{r}
AIC(m1,m2,m3)
```


EXTRA: 2.4.5. Without log transformation, poisson distribution with log link
```{r}
m5 <- bam(ENTRIESn_hourly~fog+rain+s(tempi)+s(wspdi)+as.factor(UNIT),data=df_reg.train,family=poisson(link='log'))
plot(m5)
summary(m5)
rsq5 <- rsq(df_reg.train$ENTRIESn_hourly,fitted(m5,df_reg.train))
rsq5
rsq5 <- rsq(df_reg.test$ENTRIESn_hourly,predict(m5,df_reg.test,type='response'))
rsq5
plot(df_reg.train$ENTRIESn_hourly,residuals(m5,type='deviance'))
```



EXTRA: 2.4.6 Without log transformation, Gamma distribution with inverse link
```{r}
m6 <- bam(ENTRIESn_hourly~fog+rain+s(tempi)+s(wspdi)+as.factor(UNIT),data=df_reg.train,family=Gamma(link='inverse'))
plot(m6)
summary(m6)
rsq6 <- rsq(df_reg.train$ENTRIESn_hourly,fitted(m6,df_reg.train))
rsq6
rsq6 <- rsq(df_reg.test$ENTRIESn_hourly,predict(m6,df_reg.test,type='response'))
rsq6
plot(df_reg.train$ENTRIESn_hourly,residuals(m6,type='deviance'))
```



